Fusing Filters with Integer Linear Programming

### 2014-06-22 2014betterfusionforfilters/section/03-SizeInference.tex
typesetting/indentation fixes, etc

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{alltt}
 filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
   Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}
However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$
Rule~(SProgram) also characterises the programs we accept: a program is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ program from Section~\ref{s:SizeTypes}
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}
The size constraints for this program, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}
The above program filters its input array, and then applies @map2@ to the filtered version as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

>>>>

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{alltt}
 filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
   Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}

However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & \true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  \true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

Rule~(SProgram) also characterises the programs we accept: a program is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ program from Section~\ref{s:SizeTypes}
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}

The size constraints for this program, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}

The above program filters its input array, and then applies @map2@ to the filtered version as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

### 2014-06-22 2014betterfusionforfilters/section/04-ILP.tex
rename function in CNF definition to program: should reduce confusion about function vs worker function
typesetting/indentation fixes, etc

\begin{figure}
\begin{tabbing}
MMMMM       \= M  \= \kill
$nodes$     \> @:@ \> $\function \to V$              \\
$edges$     \> @:@ \> $\function \to E$              \\
$edge$      \> @:@ \> $\{bind\} \times bind \to E$   \\
$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$   \\
\\
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\\
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,\from)$ \\
    \> $|$ \> $(\from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, \from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge \from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, \from, \fusible)$
\end{tabbing}

@@

% -----------------------------------------------------------------------------
\subsection{ILP Variables}
After generating the dependency graph, the next step is to produce a set of linear constraints from this graph. The variables involved in these constraints are split into three groups:
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{i,j}$ which indicates whether those two nodes are fused. We use $x_{i,j} = 0$ when the nodes are fused and $x_{i,j} = 1$ when they are not. Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{i,j}$ variables, and minimizing it tends to increase the number of nodes that are fused. The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{i,j} = 0 \iff cluster(i) = cluster(j)$.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic. This means that for each node in the graph, the dependencies of that node can be executed before the node itself. For each node $i$, we associate a real $\pi_i$ such that every node $j$ that depends on $i$ we have $\pi_j > \pi_i$. Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value. Here is an example of a cyclic clustering:
\begin{code}
  cycle xs  = let ys  = map (+1) xs     (C1)
                  sum = fold ys         (C2)
                  zs  = map (+sum) ys   (C1)
              in  zs
\end{code}
There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@. If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The final group of variables is used to help define the cost model encoded by the objective function. Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}. When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.

@@

% -------------------------------------
\paragraph{Acyclic and precedence-preserving} The first group of constraints ensures that the clustering is acyclic:
\begin{tabbing}
MM  \= MMMx \= M \= MMM \= M \= MMMM \= \kill
    \> ~~~~~~~ $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
    \>             (with an edge from $i$ to $j$)            \\
    \> $-N \cdot  x_{i,j}$  \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
    \>             (with no edge from $i$ to $j$)
\end{tabbing}
As per Megiddo~\cite{megiddo1998optimal} the form of these constraints is determined by whether there is an dependency between nodes $i$ and $j$. The $N$ value is set to the total number of nodes in the graph.

@@

% -------------------------------------
\paragraph{Fusion-preventing edges} As per Megiddo~\cite{megiddo1998optimal}, if there is a fusion preventing edge between two nodes we add a constraint to ensure the nodes will be placed in different clusters.
\begin{tabbing}
MMM     \= MMM \= M  \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$ \> $=$ \> $1$ \>   \> \\
        \> (for fusion-preventing edges from $i$ to $j$) 
\end{tabbing}
When combined with the precedence-preserving constraints earlier, setting $x_{i,j} = 1$ also forces $\pi_i < \pi_j$.

% -------------------------------------
\paragraph{Fusion between different iteration sizes} This group of constraints restricts which nodes can be placed in the same cluster based on their iteration size. The group has three parts. 
Firstly, either of the two nodes connected by an edge have an unknown ($\bot$) iteration size then they cannot be fused and we set $x_{i,j} = 1$:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(i) = \bot 
                ~\vee~ \iiter_{\Gamma,C}(j) = \bot$)
\end{tabbing}
Secondly, if the two nodes have different iteration sizes and no common parent then they also cannot be fused and we set $x_{i,j} = 1$:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) 
                ~\wedge~ parents(i,j) = \emptyset$)
\end{tabbing}
Finally, if the two nodes had different iteration sizes but \emph{do} have parent transducers of the same size, then the two nodes can be fused if they are fused with their respective parents, and the parents themselves are fused:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{a,A}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> $x_{b,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> $x_{A,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(a) \not= \iiter_{\Gamma,C}(b) 
                ~\wedge~ parents(a,b) = \{(A,B)\}$)
\end{tabbing}
This last part is the main difference to existing ILP solutions: we allow nodes with different iteration sizes to be fused when their parent transducers are fused. The actual constraints encode a ``no more fused than'' relationship. For example $x_{a,A} \le x_{a,b}$ means that nodes $a$ and $b$ can be no more fused than nodes $a$ and $A$.

As a simple example, consider fusing an operation on filtered data with its generating filter:
\begin{code}
    sum1 = fold (+) 0  xs
    gts  = filter (>0) xs
    sum2 = fold (+) 0  gts
\end{code}
Here $sum1$ and $sum2$ have different iteration sizes and we have that $parents(sum1, sum2) = \{(sum1, gts)\}$. This means that $sum1$ and $sum2$ may only be fused if $sum1$ is fused with $sum1$ (trivial), $sum2$ is fused with $gts$, and $sum1$ is fused with $gts$.

@@

% -------------------------------------
\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$    \> $\le$ \> $c_i$  \> \> \\
        \> (for all edges from i)
\end{tabbing}
Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory. As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.

% -----------------------------------------------------------------------------
\subsection{Objective Function}
\label{s:ObjectiveFunction}
The objective function defines the cost model of the program, and the ILP solver will find the clustering that minimizes this function while satisfying the constraints defined in the previous section. The cost model we use in this paper has three components:
\begin{itemize}
\item
the number of array reads and writes --- an abstraction of the amount of memory bandwidth needed by the program; 
\item
the number of intermediate arrays --- an abstraction of the amount of intermediate memory needed; 
\item
the number of distinct clusters --- an abstraction of the cost of loop management instructions, which maintain loop counters and the like.
\end{itemize}
The three components of the cost model are a heuristic abstraction of the true cost of executing the program on current hardware. They are ranked in order of importance --- so we prefer to minimize the number of array reads and writes over the number of intermediate arrays, and to minimize the number of intermediate arrays over the number of clusters. However, minimizing one component does not necessarily minimize any other. For example, as the fused program executes multiple array operations at the same time, in some cases the clustering that requires the least number of array reads and writes uses more intermediate arrays than strictly necessary.

@@

\begin{tabbing}
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
\\[1ex]
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible'(a,b)$ \\
    \>$=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
\end{tabbing}
With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together. Doing this reduces the total number of constraints, and makes the job of the ILP solver easier. The final formulation of the integer linear program follows.

>>>>

\begin{figure}
\begin{tabbing}
MMMMM       \= M  \= \kill
$nodes$     \> @:@ \> $\program \to V$              \\
$edges$     \> @:@ \> $\program \to E$              \\
$edge$      \> @:@ \> $\{bind\} \times bind \to E$   \\
$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$   \\
\\
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\\
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,\from)$ \\
    \> $|$ \> $(\from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, \from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge \from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, \from, \fusible)$
\end{tabbing}

@@

% -----------------------------------------------------------------------------
\subsection{ILP Variables}
After generating the dependency graph, the next step is to produce a set of linear constraints from this graph. The variables involved in these constraints are split into three groups:
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}

For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{i,j}$ which indicates whether those two nodes are fused. We use $x_{i,j} = 0$ when the nodes are fused and $x_{i,j} = 1$ when they are not. Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{i,j}$ variables, and minimizing it tends to increase the number of nodes that are fused. The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{i,j} = 0 \iff cluster(i) = cluster(j)$.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}

The second group of variables is used to ensure that the clustering is acyclic. This means that for each node in the graph, the dependencies of that node can be executed before the node itself. For each node $i$, we associate a real $\pi_i$ such that every node $j$ that depends on $i$ we have $\pi_j > \pi_i$. Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value. Here is an example of a cyclic clustering:
\begin{code}
  cycle xs  = let ys  = map (+1) xs     (C1)
                  sum = fold ys         (C2)
                  zs  = map (+sum) ys   (C1)
              in  zs
\end{code}

There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@. If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}

The final group of variables is used to help define the cost model encoded by the objective function. Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}. When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.

@@

% -------------------------------------
\paragraph{Acyclic and precedence-preserving} The first group of constraints ensures that the clustering is acyclic:
\begin{tabbing}
MM  \= MMMx \= M \= MMM \= M \= MMMM \= \kill
    \> ~~~~~~~ $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
    \>             (with an edge from $i$ to $j$)            \\
    \> $-N \cdot  x_{i,j}$  \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
    \>             (with no edge from $i$ to $j$)
\end{tabbing}

As per Megiddo~\cite{megiddo1998optimal} the form of these constraints is determined by whether there is an dependency between nodes $i$ and $j$. The $N$ value is set to the total number of nodes in the graph.

@@

% -------------------------------------
\paragraph{Fusion-preventing edges} As per Megiddo~\cite{megiddo1998optimal}, if there is a fusion preventing edge between two nodes we add a constraint to ensure the nodes will be placed in different clusters.
\begin{tabbing}
MMM     \= MMM \= M  \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$ \> $=$ \> $1$ \>   \> \\
        \> (for fusion-preventing edges from $i$ to $j$) 
\end{tabbing}

When combined with the precedence-preserving constraints earlier, setting $x_{i,j} = 1$ also forces $\pi_i < \pi_j$.

% -------------------------------------
\paragraph{Fusion between different iteration sizes} This group of constraints restricts which nodes can be placed in the same cluster based on their iteration size. The group has three parts. 
Firstly, either of the two nodes connected by an edge have an unknown ($\bot$) iteration size then they cannot be fused and we set $x_{i,j} = 1$:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(i) = \bot 
                ~\vee~ \iiter_{\Gamma,C}(j) = \bot$)
\end{tabbing}

Secondly, if the two nodes have different iteration sizes and no common parent then they also cannot be fused and we set $x_{i,j} = 1$:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) 
                ~\wedge~ parents(i,j) = \emptyset$)
\end{tabbing}

Finally, if the two nodes had different iteration sizes but \emph{do} have parent transducers of the same size, then the two nodes can be fused if they are fused with their respective parents, and the parents themselves are fused:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{a,A}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> $x_{b,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> $x_{A,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
        \> (if $\iiter_{\Gamma,C}(a) \not= \iiter_{\Gamma,C}(b) 
                ~\wedge~ parents(a,b) = \{(A,B)\}$)
\end{tabbing}

This last part is the main difference to existing ILP solutions: we allow nodes with different iteration sizes to be fused when their parent transducers are fused. The actual constraints encode a ``no more fused than'' relationship. For example $x_{a,A} \le x_{a,b}$ means that nodes $a$ and $b$ can be no more fused than nodes $a$ and $A$.

As a simple example, consider fusing an operation on filtered data with its generating filter:
\begin{code}
    sum1 = fold (+) 0  xs
    gts  = filter (>0) xs
    sum2 = fold (+) 0  gts
\end{code}

Here $sum1$ and $sum2$ have different iteration sizes and we have that $parents(sum1, sum2) = \{(sum1, gts)\}$. This means that $sum1$ and $sum2$ may only be fused if $sum1$ is fused with $sum1$ (trivial), $sum2$ is fused with $gts$, and $sum1$ is fused with $gts$.

@@

% -------------------------------------
\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$    \> $\le$ \> $c_i$  \> \> \\
        \> (for all edges from i)
\end{tabbing}

Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory. As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.

% -----------------------------------------------------------------------------
\subsection{Objective Function}
\label{s:ObjectiveFunction}
The objective function defines the cost model of the program, and the ILP solver will find the clustering that minimizes this function while satisfying the constraints defined in the previous section. The cost model we use in this paper has three components:
\begin{itemize}
\item
the number of array reads and writes --- an abstraction of the amount of memory bandwidth needed by the program; 
\item
the number of intermediate arrays --- an abstraction of the amount of intermediate memory needed; 
\item
the number of distinct clusters --- an abstraction of the cost of loop management instructions, which maintain loop counters and the like.
\end{itemize}

The three components of the cost model are a heuristic abstraction of the true cost of executing the program on current hardware. They are ranked in order of importance --- so we prefer to minimize the number of array reads and writes over the number of intermediate arrays, and to minimize the number of intermediate arrays over the number of clusters. However, minimizing one component does not necessarily minimize any other. For example, as the fused program executes multiple array operations at the same time, in some cases the clustering that requires the least number of array reads and writes uses more intermediate arrays than strictly necessary.

@@

\begin{tabbing}
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible(a,b)$ \\
    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
\\[1ex]
MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
$possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
MM        \= M    \= \kill
$possible'(a,b)$ \\
    \>$=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
\end{tabbing}

With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together. Doing this reduces the total number of constraints, and makes the job of the ILP solver easier. The final formulation of the integer linear program follows.
